Parameter Recovery Validation

### 2025-06-12 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): add dimless latent time piecewise activation model description to model progression and rename sections
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

### Dimensionless Simulated Dynamics with Observed Time

This model replaces analytical solutions with numerical integration:

- **Time coordinates**: Observed dimensionless time $t^*_j$
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$
- **Prior**: Log-normal hierarchical priors
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

### Dimensionless Simulated Dynamics with Latent Time

This model represents the full complexity of the PyroVelocity modular implementation:

- **Time coordinates**: Latent $t^*_j$ with hierarchical Gamma/Normal priors
- **Dynamics**: Numerical integration of dimensionless ODEs using torchode
- **Parameters**: $\gamma^*_i$, $U_{0i}$, $\lambda_j$, plus hierarchical time structure ($T_{M}^*$, $t_{loc}$, $t_{scl}$)
- **Prior**: Gamma priors for time scales, Normal for relative positions
- **Observation Model**: Same Poisson scaling approach
- **Likelihood**: Poisson

@@

# PyroVelocity imports
from pyrovelocity.models.modular.factory import (
    # create_simple_deterministic_model_with_validation,
    create_legacy_model2,
)
# from pyrovelocity.models.modular.components.validation import (
#     ParameterRecoveryConfig,
# )
from pyrovelocity.io.datasets import bifurcation_14

@@

## Validation Study 1: Dimensionless analytical dynamics with observed time

@@

## Validation Study 2: Dimensionless analytical dynamics with latent time

@@

The model uses the same dimensionless dynamics and observation model as Validation Study 1, but replaces observed time coordinates with **hierarchical latent time coordinates**. This approach borrows strength across cells while allowing individual temporal variation.

@@

## Validation Study 3: Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation

@@

where:
- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
- $\delta^*_i$ is the dimensionless activation duration

@@

In the PGM above, we use simplified notation for visual clarity:
- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- **Key edge**: $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters

>>>>

### Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation

This model extends the latent time framework with piecewise constant transcription rates:

- **Time coordinates**: Same hierarchical latent $t^*_j$ structure as previous model
- **Dynamics**: Piecewise constant transcription with three phases (off, on, return-to-off)
- **Parameters**: $\gamma^*_i$, $R_{\text{on},i}$ (fold-change), $t^*_{0,\text{on}i}$ (onset), $\delta^*_i$ (duration), plus hierarchical time structure
- **Key innovation**: Fixed reference state ($\alpha^*_{\text{off}i} = 1.0$) eliminates parameter redundancy
- **Enhanced patterns**: Supports pre-activation scenarios with negative onset times
- **Analytical solutions**: Closed-form solutions for all three phases, including $\gamma^*_i = 1$ special case
- **Biological realism**: Steady-state initial conditions and rich expression pattern representation
- **Observation Model**: Same Poisson scaling with characteristic scales and capture efficiency
- **Likelihood**: Poisson

@@

# PyroVelocity imports
from pyrovelocity.models.modular.factory import (
    create_piecewise_activation_model,
    create_legacy_model2,
)
# from pyrovelocity.models.modular.components.validation import (
#     ParameterRecoveryConfig,
# )
from pyrovelocity.io.datasets import bifurcation_14

@@

## Dimensionless analytical dynamics with observed time {#sec-dimless-observed-time}

@@

## Dimensionless analytical dynamics with latent time {#sec-dimless-latent-time}

@@

The model uses the same dimensionless dynamics and observation model as the model described in @sec-dimless-observed-time, but replaces observed time coordinates with **hierarchical latent time coordinates**. This approach borrows strength across cells while allowing individual temporal variation.

@@

## Dimensionless Analytical Dynamics with Latent Time and Piecewise Activation {#sec-dimless-latent-piecewise}

@@

- $\alpha^*_{\text{off}i} = 1.0$ is the fixed reference transcription rate
- $R_{\text{on},i} = \alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} = \alpha^*_{\text{on}i}$ is the fold-change during activation (inferred parameter)
- $t^*_{0,\text{on}i}$ is the dimensionless activation onset time (can be negative for pre-activation scenarios)
- $\delta^*_i$ is the dimensionless activation duration

@@

In the PGM above, we use simplified notation for visual clarity:

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- **Key edge**: $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters
